The human in the loop: volunteer computing as a socio-technical system

### 2016-02-28 Alife16/paper.tex
Reduced state of the art to the relevant parts.
Refs #101 but I still have to update it.

The web used as a resource for distributed computing has a
long history, and almost since the beginning it was used for
evolutionary computation even before JavaScript was
introduced as a browser-based scripting language in 1997. Before that
time other technologies,
 including Flash animations and VBScript, ActiveX or Java applets were
used; for instance, Java was pointed out in \citep{soares1998get} as a
``language for the
internet'', providing some advantages such as multi-architecture compatibility or
security mechanisms, and was proposed in \citep{chandy1996world} as the
basis for the creation of a world-wide peer to peer system using compiled Java
programs. The ATLAS system
\citep{Baldeschwieler:1996:TIG:504450.504482} also used Java together
with another language called Cilk; all these papers indicated that
the introduction of an architecture-independent system such as the
Java virtual machine opened the doors for ultrascale, Internet-wide,
metacomputers. This paper also establishes the desirable features of
any {\em global computing} infrastructure, that can also be applied to
distributed evolutionary computation experiments that rely on
volunteer computing: scalability, heterogeneity, fault tolerance,
adaptive parallelism, safety, anonymity, hierarchy, ease of use, and
reasonable performance. But hierarchy, this is, the capability of
preferentially using resources in the same organization, is a feature
that could be either disregarded or included within adaptivity, which
leaves us with eight desirable properties.

The fulfillment of these properties could be achieved by using any of the
languages and associated technologies, mentioned above, provided that the user had
installed the required virtual machine plugin first.

However, embedding a
distributed computation client in the browser has the advantage of
providing a single and universal way of accessing resources via an
URI, something that is not guaranteed by using desktop applications. 
Baratloo et al. tried in 1996 to create a framework with
the aforementioned qualities called {\em Charlotte}
\citep{baratloo1996charlotte}. This framework, which implemented a
distributed shared memory, was inspired by a previous work
on parallel computing systems PVM and MPI and solved most of the issues by
using the Java virtual machine embedded in the browser,
which provided
security to the user, uniformity of computing environment
independently of the operating system, and access to a common data store in the
server. Since the Java virtual machine is a sandbox, it is safe to the user
and as it runs compiled bytecode it also guarantees a certain level of
performance.
Other issues such as scheduling, which are related to adaptive 
parallelism, were solved on the server. This
paper was later extended and published in a journal in 1999
\citep{baratloo1999charlotte}, but it is interesting to note that,
even in the early stages of the web, the possibility of
browser-based distributed computing was already researched.

Other systems, like JET, proposed by Soares et
al. \citep{soares1998get}, also used Java. JET supports
the execution of parallel applications over the Internet and features
a comprehensive statistics collection subsystem, enabling its use for science, 
since it allows 
the collection of algorithm analytics by the
experiment designer, unlike the other systems mentioned
above. Other proposals contemporary to JET, such as {\em SuperWeb}
\citep{alexandrov1997superweb} are roughly a translation of classical
distribution techniques, with overly complicated architectures that include
brokers for resource discovery. SuperWeb is focused,
however, in solving the
issues involved in browser-based computing and also in making a system
that is able to work across different types of computers, mentioning
SPARCstations and PCs with Windows-95. Interestingly enough, this
paper incorporates a discussion on the economic model, paving the way
for the for-profit distributed computing systems that arose later on. 
On the other hand, the potential for volunteer computing using
browsers was realized
later on \citep{sarmenta-bayanihan} as well as the potential of
user misconduct when having access to the source code.
\citep{sarmenta-sabotagetolerance}. However, these early efforts by
Sarmenta once again used Java and not JavaScript, making them % (Paloma) making them
less widespread.

The main problem with Java and all the other browser-embedded
technologies is that they are not {\em universal} in the
sense that an extra component, namely, the Java virtual machine or another
virtual environment has to be installed in the browser.
Also these virtual environments have become targets of malicious attacks; 
for instance numerous vulnerabilities have been discovered in 
the proprietary Adobe Flash runtime environment and
video player \citep{ford2009analyzing,watanabe2010new}. The need for
these additional environments has been mitigated with the
adoption of the HTML5 standard \citep{anthes2012html5}.
JavaScript \citep{flanagan2006javascript}, however, has become a set of standards
\citep{ECMA-262} that include the language and its components. Since
early on, it was adopted by the industry and also by scientists,
who used them for creating an evolutionary algorithm (EA) on
the browser as early as 1998 \citep{jj-ppsn98}.

@@

Recent works propose the use of cloud computing services for the components of
client-server distributed EAs. Cloud resources
can reduce operational costs by outsourcing both hardware and software maintenance
to the cloud provider. Another advantage is that they enable the provisioning of computing resources beyond what
is available in most laboratories, allowing researches to
scale algorithms at reasonable costs. Researchers have proposed the use of
public clouds such as Amazon EC2 \citep{CloudScale}, Google App Engine\citep{di2013towards},
and DropBox~\citep{mericloud}. The use of JavaScript to implement
server side components
gives software developers certain advantages in the cloud
paradigm, since it is a first-class citizen
in most Platforms as a Service (PaaS) products as the {\tt Node.js} framework is supported by most of them \citep{wood13:nodejs:paas}. Many web services
have native JavaScript libraries too, and also use the JSON (JavaScript Object Notation) format for data interchange. Another advantage is that
developers can use the same software tools for both client and server
side development.

JavaScript is also compatible with the functional programming paradigm \citep{Cousineau1998,MacLennan1990,Thompson1996}.
According to \citep{swanresearch2015}, the benefit of using this kind of paradigm in
metaheuristics optimization is that it allows an improvement in communicability,
reproductibility, interoperability, automated assembly, knowledge engineering
and efficiency.

In the work presented in this paper, JavaScript has been used throughout the full
stack, which has the advantage of using the self-same lines of code for
writing an evolutionary algorithm for the
desktop or command line and browser clients, both based in the EC
library {\sf NodEO} \citep{DBLP:conf/gecco/GuervosVGES14}. This saves time 
when writing code and leverages 
the expertise acquired in the library itself.

On the other hand, we are also interested in measuring the performance
of volunteer computing systems, an area in which there have been
relatively few efforts.
There were some initial attempts to avoid the differences in performance
that could be obtained from volunteers  by making
the algorithm adaptive to the kind of resources allotted to it
\citep{milani2004online}, which is actually not such a big problem in
algorithms such as the EA that can easily be
parallelized via population splitting or farming out evaluations to all
the nodes available. Lately, several approaches have focused on the
fault-tolerance of volunteer algorithms
\citep{gonzalez2010characterizing} which it can, of course, be studied in
the more general context of distributed computing 
\citep{nogueras2015studying} or including it in a more general study of the
performance of the EA itself
\citep{DBLP:journals/gpem/LaredoBGVAGF14}.

On the other hand, measuring the performance of the resulting metacomputer
involves understanding the dynamics of this kind of systems. Initial
work was done for peer to peer systems by Stutzbach et
al. \citep{stutzbach2006understanding} and extended to volunteer
computing by Laredo et al. \citep{churn08,laredo2008rcp}.

But some of the essential metrics in volunteer computing like the
number of users or the time spent by every one in the
computation in browser-based volunteer computing experiments, have
only been studied in a limited way in 
\citep{DBLP:journals/gpem/LaredoBGVAGF14} on the basis of a single
run. Studies using volunteer computing platforms such as SETI@home
\citep{javadi2009mining} found out that the Weibull, log-normal and
Gamma distribution 
modeled quite well the availability of resources in several clusters
of that framework; the shape of those distributions is a skewed bell
with more resources in the {\em low} areas than in the high areas:
there are many users that give a small amount of cycles, while there
are just a few that give many cycles. This is in concordance with the
results obtained in \citep{agajaj}. Although SETI@home works
at a larger scale in both quantity of volunteers and the scope of projects, 
our initial tests indicate that
this kind of distribution might be the most appropriate to describe
user participation in several different metrics. However, in this
paper we will perform different experiments in different setups and
will try to come up with a more precise model. Next we describe the
framework itself.

As far as we know, this paper presents one of the few experiments using 
computational resources that are as dissimilar as smartphones and
powerful laptops 
or workstation computers in a research center. 
%The methodology used
% (Paloma) I would say "the used methodology"
The algorithms used, as well as the methodology 
for gathering resources will be described next, 
together with the results obtained in this initial setup.

@@

In general, a distributed volunteer-based evolutionary computation
system based on the browser is simply a client-server system
whose client is, or it can be, embedded in the browser via
JavaScript. Since JavaScript is  the only language that is present
across all browsers, the choice was quite clear. Even so, there might
be some issues with the performance of the language itself; which % (Paloma) I recommend the use of "which" here
 is
why we have made a comparison between JavaScript and other languages
\citep{2015arXiv151101088M}. The results of this work show that JavaScript can be
faster than compiled languages such as Scala and, in any case, it provides
a performance that is comparable to other languages usually employed % (Paloma) Adverb
in evolutionary computation and numerical optimization such as Python. 
% (Paloma) Bit of a too long sentence here. Why not changing the ; for a . ?
Compiled languages such as
Java or C might be faster when % (Paloma) Too many "fact" words. There was another one in the previous sentence.
considering the performance of a single-user, 
but this is more than offset by the computational power of
the spontaneous volunteers we can gather at whim, including people
using their mobile phones or tablets. Together, the performance is several orders of magnitude
higher, which is the objective in this kind of systems.
% (Paloma) This sentence needs re-writing
% "Other compiled languages such as Java or C might be, in fact, faster, but the fact that single-user performance is higher (than what?) is over the offset. This is because the number of spontaneous volunteers we can gather at whim, including people using their mobile phones or tablets, is several orders of magnitude higher, which is the objective in this kind of systems." (?)
 JavaScript is
quite a popular language nowadays, providing compilers from a number
of languages, many of them in their family (like CoffeeScript), but
many other outside, so researchers can, in fact, write their fitness
function in Python, Erlang, Scala, or even Java, and them compile it to
JavaScript. For completeness, an experiment comparing the performance of
a JavaScript implementation versus Java and Matlab is included in section
\ref{sec:w2}.

>>>>

Volunteer computing involves an user running a program voluntarily
and, as such, has been deployed in many different ways from the
beginning of the Internet, starting with the SETI@home framework for
processing extraterrestrial signals \cite{david-seti:home}. However
the dual introduction of JavaScript as an universal language for the
browser and the browser as an ubiquitous web and Internet client has
made this combination the most popular for volunteer computing
frameworks such as the one we are using here and whose first version
was described in \cite{DBLP:conf/gecco/GuervosG15}.

@@

On the other hand, we are also interested in measuring the performance
of volunteer computing systems, an area in which there have been
relatively few efforts.
There were some initial attempts to avoid the differences in performance
that could be obtained from volunteers  by making
the algorithm adaptive to the kind of resources allotted to it
\citep{milani2004online}, which is actually not such a big problem in
algorithms such as the EA that can easily be
parallelized via population splitting or farming out evaluations to all
the nodes available. Lately, several approaches have focused on the
fault-tolerance of volunteer algorithms
\citep{gonzalez2010characterizing} which it can, of course, be studied in
the more general context of distributed computing 
\citep{nogueras2015studying} or including it in a more general study of the
performance of the EA itself
\citep{DBLP:journals/gpem/LaredoBGVAGF14}. This performance cannot be
measured without understanding the dynamics of this kind of systems. Initial
work was done for peer to peer systems by Stutzbach et
al. \citep{stutzbach2006understanding} and extended to volunteer
computing by Laredo et al. \citep{churn08,laredo2008rcp}.

Some of the essential metrics in volunteer computing like the
number of users or the time spent by every one in the
computation in browser-based volunteer computing experiments, have
only been studied in a limited way in 
\citep{DBLP:journals/gpem/LaredoBGVAGF14} on the basis of a single
run. Studies using volunteer computing platforms such as SETI@home
\citep{javadi2009mining,agajaj} found out that the Weibull, log-normal and
Gamma distribution 
modeled quite well the availability of resources in several clusters
of that framework; the shape of those distributions is a skewed bell
with more resources in the {\em low} areas than in the high areas:
there are many users that give a small amount of cycles, while there
are just a few that give many cycles.

As far as we know, this paper presents one of the few experiments
that measure the performance of a sociotechnical metacomputer. 
The algorithms used, as well as the methodology 
for gathering resources will be described next, 
together with the results obtained in this initial setup.

@@

In general, a distributed volunteer-based evolutionary computation
system based on the browser is simply a client-server system
whose client is, or it can be, embedded in the browser via
JavaScript. Since JavaScript is  the only language that is present
across all browsers, the choice was quite clear. Even so, there might
be some issues with the performance of the language itself; which % (Paloma) I recommend the use of "which" here
 is
why we have made a comparison between JavaScript and other languages
\citep{2015arXiv151101088M}. The results of this work show that JavaScript can be
faster than compiled languages such as Scala and, in any case, it provides
a performance that is comparable to other languages usually employed % (Paloma) Adverb
in evolutionary computation and numerical optimization such as Python. 
% (Paloma) Bit of a too long sentence here. Why not changing the ; for a . ?
Compiled languages such as
Java or C might be faster when % (Paloma) Too many "fact" words. There was another one in the previous sentence.
considering the performance of a single-user, 
but this is more than offset by the computational power of
the spontaneous volunteers we can gather at whim, including people
using their mobile phones or tablets. Together, the performance is several orders of magnitude
higher, which is the objective in this kind of systems.
% (Paloma) This sentence needs re-writing
% "Other compiled languages such as Java or C might be, in fact, faster, but the fact that single-user performance is higher (than what?) is over the offset. This is because the number of spontaneous volunteers we can gather at whim, including people using their mobile phones or tablets, is several orders of magnitude higher, which is the objective in this kind of systems." (?)
 JavaScript is
quite a popular language nowadays, providing compilers from a number
of languages, many of them in their family (like CoffeeScript), but
many other outside, so researchers can, in fact, write their fitness
function in Python, Erlang, Scala, or even Java, and them compile it to
JavaScript.
